---
title: Demo of Confidence Functions for Intro Stat Topics
date: 090820
author: David Darmon
---

```{r, message = FALSE}
# devtools::install_github('ddarmon/clp') # Only need to run this once, to install clp
library(clp)
```

# Single Gaussian Mean
## From summary data

```{r}

m1 <- t.conf.summary(mean = 1, sd = 2, n = 5)

m1$qconf(c(0.025, 0.975)) # 95% confidence interval

# Tests for H0: mu = 0

m1$pconf(0) # Right-sided P-value
1-m1$pconf(0) # Left-sided P-value

m1$pcurve(0) # Two-sided P-value
```
## From actual data

```{r}
data(dietstudy)

m1 <- t.conf(dietstudy$weightchange[dietstudy$diet == 'Low Carb'], conf.level = 0.999)
```

# Paired Means

```{r}
data(sleep)

mp <- t.conf(sleep$extra[sleep$group == 1],
            sleep$extra[sleep$group == 2], paired = TRUE)

1-mp$pconf(0) # P-value for H0: mu >= 0
```
# Welch's Two-sample T-test

```{r}
# Women's weights
x <- c(38.9, 61.2, 73.3, 21.8, 63.4, 64.6, 48.4, 48.8, 48.5, 43.6)

# Men's weights
y <- c(67.8, 60, 63.4, 76, 89.4, 73.3, 67.3, 61.3, 62.4, 111.2) 

m2s <- t.conf(x, y)

m2s$pcurve(0) # P-value for no difference in mean weights between men and women
```

# Test of one proportion

```{r}
p1 <- prop.conf(x = 1, n = 10)

p1$qconf(c(0.01, 0.99)) # 98% confidence interval for p
```
# Test of two proportions (difference)

```{r}
p2s <- prop.conf(x = c(15, 5), n = c(30, 30))

p2s$pcurve(0) # Test for p1 - p2 = delta = 0
```

# Test of two proportions (risk)

```{r}
riskratio <- riskratio.conf(x = c(15, 5), n = c(30, 30))

riskratio$pcurve(1) # Test for no difference in risk

riskratio <- riskratio.conf(x = c(15, 5), n = c(30, 30), log = 'x') # Log-scale for confidence curve
```
# Test of two proportions (odds)

```{r}
odds <- oddsratio.conf(x = c(15, 5), n = c(30, 30))

odds$pcurve(1) # Test for no difference in odds

odds <- oddsratio.conf(x = c(15, 5), n = c(30, 30), log = 'x') # Log-scale for confidence curve
```
# Linear Regression with Gaussian Noise Assumption

```{r}
data(faithful)

mod <- lm(eruptions ~ waiting, data = faithful)

conf.lm <- lm.beta.conf(mod)

display.cconf(conf.lm$`(Intercept)`, xlab = '(Intercept)')
display.cconf(conf.lm$waiting, xlab = 'waiting')
```
# Correlation

```{r}
corr.out <- cor.conf(faithful$eruptions, faithful$waiting)
```


# Wilcoxon's signed-rank test

```{r}
d <- sleep$extra[sleep$group == 1] - sleep$extra[sleep$group == 2]

mp <- wilcox.conf(d)

1-mp$pconf(0) # P-value for median >= 0
```
# Wilcoxon's rank-sum test

```{r}
# Women's weights
x <- c(38.9, 61.2, 73.3, 21.8, 63.4, 64.6, 48.4, 48.8, 48.5, 43.6)

# Men's weights
y <- c(67.8, 60, 63.4, 76, 89.4, 73.3, 67.3, 61.3, 62.4, 111.2) 

m2s <- wilcox.conf(x, y)

m2s$pcurve(0) # P-value for no difference in shift between mean and women
```
# Multinomial

```{r}
N <- c(315, 108, 101, 32)
names(N) <- c('Round, Yellow', 'Round, Green', 'Wrinkled, Yellow', 'Wrinkled, Green')

theta0 <- c(9, 3, 3, 1)/16

col <- c('darkgoldenrod', 'darkgreen', 'darkgoldenrod1', 'darkolivegreen')

B <- 2000

conf.out <- multinomial.conf(N, B = B, col = col)
abline(v = theta0, col = col, lty = 3)

# Combine evidence across confidence curves.
p.multinomial(conf.out, theta0)

chisq.test(N, p = theta0)
```
# Chi-squared test of association

```{r}
# Galton's data on assortative mating according to temper.

# Columns: wife (good / bad) temper
# Rows:    husband (good / bad) temper

# reported on page 78 of *Confidence, Likelihood, Probability*.

N <- matrix(c(24, 27,
              34, 26), nrow = 2, byrow = TRUE)

conf.out <- chisq.conf(N)
```

